---
title: "NEON Data Analysis of Verrucomicrobiota and Niwot Ridge"
subtitle: "Evo-Geno Final Report Spring 2024"
author: "Michelle Le, Olivia Mooradian, Jason Chen, Sofia Blomqvist"
output: 
  html_document:
    toc: true
    toc_depth: 4
    theme: cerulean
    highlight: default
    code_folding: hide
bibliography: NEON.bib
---
# Abstract

Verrucomicrobiota are present in almost all soil, but it was originally thought to be at low frequencies since there were only a few studies that focused on the Verrucomicrobiota quantities. However, with ecological data from The National Ecological Observatory Network (NEON), funded by the National Science Foundation, more data has surfaced, providing information about the bacteria across ecosystems, with our data focusing on Niwot Ridge, Colorado. With the increasing temperatures due to global warming, it is important to take notice of the different bacteria that are growing and changing along with the climate. Where they require very specific environments to grow, taking notice of the different features of existing environments is massively important. In the near future where climates might not be able to keep the bacteria alive, it would be vital to provide a possible man made environment that can facilitate the bacteria growth with the data collected, allowing for a deeper understanding of the bacteria.

# Motivating Reasons

Bacteria play many important roles in the soil of ecosystem such as providing soil structure, soil aggregation, recycling of soil nutrients, and water recycling. They also serve an important function in breaking down organic residues, carbon compounds and simple sugars, as well as taking up nutrients into their cell membranes such as nitrogen [@noauthor_role_nodate]. These functions are important for plant growth and maintaining a healthy ecosystem, i.e. continuous recycling of nutrients, maintaining a balance of pathogenic and non-pathogenic bacterial species, and soil stability. The phylum of bacteria Verrucromicrobiota are widespread and while the research surrounding their functions is limited it is thought that they play a major role in carbon degeneration and have a greater ability to survive in a variety of conditions than previously thought. Just like other bacteria, Verrucromicrobiota must be able to adapt to the diverse microenvironments of the soil in order to provide these ecological benefits, and with increasing global temperatures that generally raise microbial activity and in turn decrease the amount of organic matter in soil [@ruan_elevated_2023] it will be important to monitor their abundance in different ecosystems across time. 

# Introduction

The National Ecological Observatory Network (NEON) funded by the National Science Foundation is a large, continental-scale project with the purpose of collecting long-term open access ecological data at various sites throughout the United States including Hawaii and Puerto Rico. This data enables researchers to conduct studies addressing questions related to important environmental challenges. The NEON database includes information on freshwater ecosystems such as rivers and streams as well as terrestrial ecosystems ranging from mountainous tundra to tropical rainforests. We are particularly interested in the microbe metagenome sequence data that NEON has collected which includes metagenome-assembled genomes (MAGs) of various phyla and orders of bacteria at different sites across the United States [@noauthor_niwot_nodate]. For our specific project, we focus on the NEON project site located in the Niwot Ridge in Boulder, Colorado (NIWO), and the bacteria Verrucomicrobiota. In our study, we are using NEON data from the NIWO site to explore the relationship between Verrucomicrobiota abundance and soil conditions, and we explore how this relationship impacts Verrucomicrobiota growth and what this could mean for the ecosystems the bacteria inhabits.

Verrucomicrobiota is a free living, widespread phylum of bacteria surviving in freshwater, marine, terrestrial soil, and animal gut, and are significant contributors to carbon cycling. Despite their prevalence, they are underrepresented in databases, limiting our understanding of their ecological roles [@bergmann_under-recognized_2011]. Recent findings highlight Verrucomicrobiota's role as specialist organic polymer degraders capable of adapting to diverse environmental conditions, expanding our understanding of their ecological significance [@nixon_genome-resolved_2019]. 


# Methods

We imported soil microbe metagenomic sequence data from the NEON data portal into R studio. We utilized R functions in order to manipulate and analyze the data, including ggplot and ggtree. We filtered the data so that it would include only information from our site, NIWO. We differentiated the phylums and orders of bacteria by looking at metagenome-assembled genomes (MAGs) and could filter the data to only include phylum Verrucomicrobia when necessary. We then graphed the data and created 50 visuals that explored relationships between Verrucromicorbiota MAG count and soil temperature; Verrucromicorbiota MAG count and soil pH; Verrucromicrobiota MAG count and soil content; Verrucromicrobiota MAG count and soil depth, as well as how their abundance compares to other phylums of bacteria depending on soil conditions. 

### Loading Libraries and Data

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(tidyverse)
library(knitr)
library(ggtree)
library(TDbook)
library(ggimage)
library(rphylopic)
library(treeio)
library(tidytree)
library(ape)
library(TreeTools)
library(phytools)
library(ggnewscale)
library(ggtreeExtra)
library(ggstar)
library(zoo)
```

```{r, warning=FALSE, message=FALSE}
NEON_MAGs <- read_csv("data/NEON/GOLD_Study_ID_Gs0161344_NEON_2024_4_21.csv") %>% 
  select(-c(`GOLD Study ID`, `Bin Methods`, `Created By`, `Date Added`, `Bin Lineage`)) %>% 
  mutate("Assembly Type" = case_when(`Genome Name` == "NEON combined assembly" ~ `Genome Name`,
                            TRUE ~ "Individual")) %>% 
  mutate_at("Assembly Type", str_replace, "NEON combined assembly", "Combined") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "d__", "") %>%  
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "p__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "c__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "o__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "f__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "g__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "s__", "") %>%
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";", remove = FALSE) %>% 
  mutate_at("Domain", na_if,"") %>% 
  mutate_at("Phylum", na_if,"") %>% 
  mutate_at("Class", na_if,"") %>% 
  mutate_at("Order", na_if,"") %>% 
  mutate_at("Family", na_if,"") %>% 
  mutate_at("Genus", na_if,"") %>% 
  mutate_at("Species", na_if,"") %>% 

  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

```{r, warning=FALSE, message=FALSE}
NEON_metagenomes <- read_tsv("data/NEON/exported_img_data_Gs0161344_NEON.tsv") %>% 
  select(-c(`Domain`, `Sequencing Status`, `Sequencing Center`)) %>% 
  rename(`Genome Name` = `Genome Name / Sample Name`) %>% 
  filter(str_detect(`Genome Name`, 're-annotation', negate = T)) %>% 
  filter(str_detect(`Genome Name`, 'WREF plot', negate = T))
```

```{r, warning=FALSE, message=FALSE}
NEON_metagenomes <- NEON_metagenomes %>% 
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

```{r, warning=FALSE, message=FALSE}
NEON_chemistry <- read_tsv("data/NEON/neon_plot_soilChem1_metadata.tsv") %>% 
  mutate_at("genomicsSampleID", str_replace, "-COMP", "") 
```

### Creating Data Frames/Preparing Data

#### NEON

```{r}
NEON_data <- NEON_MAGs %>% 
  left_join(NEON_metagenomes, by = "Sample Name") %>% 
  left_join(NEON_chemistry, by = c("Sample Name" = "genomicsSampleID")) %>% 
  rename("label" = "Bin ID")

NEON_data_Verruco <- NEON_data %>% 
 filter(str_detect(Phylum, "Verrucomicrobiota"))

NEON_data_NIWO <- NEON_data %>% 
  filter(str_detect(Site.x, "Niwot Ridge"))
```

#### Trees

```{r}
tree_arc <- read.tree("data/NEON/gtdbtk.ar53.decorated.tree")
tree_bac <- read.tree("data/NEON/gtdbtk.bac120.decorated.tree")
```

```{r}
node_vector_bac = c(tree_bac$tip.label,tree_bac$node.label)

grep("Verrucomicrobiota", node_vector_bac, value = TRUE)
match(grep("Verrucomicrobiota", node_vector_bac, value = TRUE), node_vector_bac)

```

```{r}
tree_bac_preorder <- Preorder(tree_bac)
tree_Verruco <- Subtree(tree_bac_preorder, 2563)
```

# Results

## Environmental Data

### Soil Data - Verrucomicrobiota vs. all MAGs

#### Verrucomicrobiota Soil Temperature Distribution

```{r, warning=FALSE, message=FALSE, fig_width=10}
NEON_data_Verruco %>% 
  ggplot(aes(x = soilTemp)) +
  geom_bar()+
  labs(title = "Distribution of MAGs based on soil temperature", y = "Verrucomicrobiota MAGs", x= "Soil Temperature")
```

#### Verrucomicrobiota Soil pH Distribution

```{r, warning=FALSE, message=FALSE, fig_width=10}
NEON_data_Verruco %>% 
  ggplot(aes(x = soilInWaterpH)) +
  geom_bar()+
  labs(title = "Distribution of MAGs based on soil pH", y = "Verrucomicrobiota MAGs", x= "Soil pH")

```

#### Total MAGs Soil Temperature

```{r, warning=FALSE, message=FALSE, fig_width=10}
NEON_data %>% 
  ggplot(aes(x = soilTemp, fill= Phylum)) +
  geom_bar()+
  labs(title = "Distribution of Total MAGs based on soil temperature", y = "MAGs", x= "Soil Temperature")
```

#### Total MAGs Soil pH

```{r, warning=FALSE, message=FALSE, fig_width=12, fig_height=6}
NEON_data %>% 
  ggplot(aes(x = soilInWaterpH, fill= Phylum)) +
  geom_bar()
  labs(title = "Distribution of Total MAGs based on soil in water pH", y = "MAGs", x= "Soil pH")
```

### Ecosystems - Verrucomicrobiota vs. all MAGs

#### Verrucomicrobiota Ecosystem Distribution

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Ecosystem Subtype`))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Ecosystem Subtypes", y="Count")
```

#### Verrucomicrobiota Ecosystem Distribution by Order

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x=`Ecosystem Subtype`, fill=Order))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Distribution by Ecosystem", y="Verrucomicrobiota MAGs")
```

### Verrucomicrobiota Ecosystem Subtypes by Site

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Ecosystem Subtype`, fill = `Site ID.x`))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Ecosystem Subtype", y="Count")
```

#### Total MAGs Ecosystem Distribution

```{r, fig.height= 5, fig.width= 10}
NEON_data %>% 
  ggplot(aes(x=`Ecosystem Subtype`, fill= Phylum))+
  geom_bar()+
  labs(title = "MAG Distribution by Ecosystem", y="MAGs")
```

#### Verrucomicrobiota and Ecosystem Subtype Phylogeny

```{r, fig.height=5, fig.width=7}
ggtree(tree_Verruco, layout="circular")  %<+%
  NEON_data + 
  geom_point2(mapping=aes(color=`Ecosystem Subtype`))+
  ggtitle("Verrucomicrobiota with Ecosystem Subtypes")
```

### Other Soil Properties - Verrucomicrobiota

#### Verrucomicrobiota Distribution and Nitrogen Content

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x= nitrogenPercent, fill=Order))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Distribution by Nitrogen Content", x="Nitrogen Percent", y="Verrucomicrobiota MAGs")
```

#### Verrucomicrobiota Distribution and Organic Carbon Content

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x= organicCPercent, fill=Order))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Distribution by Organic Carbon Content", x="Organic Carbon Percent", y="Verrucomicrobiota MAGs")
```

#### Verrucomicrobiota Distribution and Depth of Sample

```{r, warning=FALSE, message=FALSE}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Depth In Meters`, fill=Order))+
  geom_bar()+
  labs(title = "Verrucomicrobiota Distribution by Depth of Sample", x="Depth (meters)", y="Verrucomicrobiota MAGs")
```

### Site Data

#### Soil Temperature and Elevation at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  ggplot(aes(x=elevation, y=soilTemp))+
  geom_point()+
  labs(title = "Relationship between soil temperature and elevation at Niwot Ridge", y = "Soil Temperature", x= "Elevation (m)")
```

#### Soil Temperature and Elevation at all sites

```{r, warning=FALSE, message=FALSE}
NEON_data %>% 
  ggplot(aes(x=elevation, y=soilTemp))+
  geom_point()+
  labs(title = "Relationship between soil temperature and elevation at all sites", y = "Soil Temperature", x= "Elevation (m)")
```

#### Soil pH and Soil Temperature - All Sites

```{r, warning=FALSE, message=FALSE}
NEON_data %>% 
  ggplot(aes(x=soilTemp, y=soilInWaterpH, color = Phylum))+
  geom_point()+
  labs(title = "Relationship between temperature and pH", y = "Soil pH", x= "Soil Temperature")
```

#### Ecosystem Subtype and Soil Temperature

```{r, warning=FALSE, message=FALSE}
NEON_data %>% 
  ggplot(aes(x=`Ecosystem Subtype`, y=soilTemp))+
  geom_boxplot()+
  labs(title = "Ecosystem Subtype Temperature Distribution", y="Soil Temperature")
```

#### Habitat of each Phylum at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  ggplot(aes(x= `Habitat`, fill=Phylum))+
  geom_bar()+
  labs(title = "Habitat of Samples at NIWO", y="Count")
```

#### Mineral vs Organic Soil at Niwot Ridge

```{r}
NEON_data_NIWO%>%
  ggplot(aes(x = Layer.x))+
  geom_bar()+
  labs(title = "Abundance of Verrucomicrobiota in Mineral vs. Organic Soil")+
  labs(x = "Mineral vs. Organic Soil", y = "Verrucomicrobiota Count")
```

#### Elevation of Samples Found at Niwot Ridge

```{r, warning=FALSE, message=FALSE}
NEON_data_NIWO %>%   
ggplot(aes(x = `elevation`, y = `Sample Name`, color = `Phylum`)) +
  geom_point() +
  ggtitle("Elevation and Samples Found at NIWO")+
  theme(axis.text.x = element_text(angle=45, vjust=1, hjust=1))
```

## Genome/Assembly Data

### Sankey Plots

```{r, message=FALSE, warning=FALSE}
NEON_MAGs_test <- read_csv("data/NEON/GOLD_Study_ID_Gs0161344_NEON_2024_4_21.csv") %>% 
  # remove columns that are not needed for data analysis
  select(-c(`GOLD Study ID`, `Bin Methods`, `Created By`, `Date Added`, `Bin Lineage`)) %>% 
  # create a new column with the Assembly Type
  mutate("Assembly Type" = case_when(`Genome Name` == "NEON combined assembly" ~ `Genome Name`,
                            TRUE ~ "Individual")) %>% 
  mutate_at("Assembly Type", str_replace, "NEON combined assembly", "Combined") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "d__", "") %>%  
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "p__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "c__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "o__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "f__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "g__", "") %>% 
  mutate_at("GTDB-Tk Taxonomy Lineage", str_replace, "s__", "") %>%
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";", remove = FALSE) %>% 
  mutate_at("Domain", na_if,"") %>% 
  mutate_at("Phylum", na_if,"") %>% 
  mutate_at("Class", na_if,"") %>% 
  mutate_at("Order", na_if,"") %>% 
  mutate_at("Family", na_if,"") %>% 
  mutate_at("Genus", na_if,"") %>% 
  mutate_at("Species", na_if,"") %>% 
  
  # Get rid of the the common string "Soil microbial communities from "
  mutate_at("Genome Name", str_replace, "Terrestrial soil microbial communities from ", "") %>% 
  # Use the first `-` to split the column in two
  separate(`Genome Name`, c("Site","Sample Name"), " - ") %>% 
  # Get rid of the the common string "S-comp-1"
  mutate_at("Sample Name", str_replace, "-comp-1", "") %>%
  # separate the Sample Name into Site ID and plot info
  separate(`Sample Name`, c("Site ID","subplot.layer.date"), "_", remove = FALSE,) %>% 
  # separate the plot info into 3 columns
  separate(`subplot.layer.date`, c("Subplot", "Layer", "Date"), "-") 
```

```{r}
NEON_MAGs_ind <- NEON_MAGs %>% 
  filter(`Assembly Type` == "Individual") 

NEON_MAGs_co <- NEON_MAGs %>% 
  filter(`Assembly Type` == "Combined") 
```

#### All Combined

```{r}
sankey_data2 <- NEON_MAGs_co %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") 


sankey_data2[] <- t(apply(sankey_data2, 1, zoo::na.locf))


sankey_data2 <- sankey_data2 %>% 
    unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")


sankey_data_s <- sankey_data2
sankey_data_g <- sankey_data2
sankey_data_f <- sankey_data2
sankey_data_o <- sankey_data2
sankey_data_c <- sankey_data2
sankey_data_p <- sankey_data2
sankey_data_d <- sankey_data2

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)   

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 

  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

write_tsv(sankey_data_allTaxa, "data/NEON/all_co_pavian.txt")
```

```{r}
knitr::include_url("images/sankey_all_co.txt.html")
```
#### All Individual

```{r}
sankey_data2 <- NEON_MAGs_ind %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") 


sankey_data2[] <- t(apply(sankey_data2, 1, zoo::na.locf))


sankey_data2 <- sankey_data2 %>% 
    unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")


sankey_data_s <- sankey_data2
sankey_data_g <- sankey_data2
sankey_data_f <- sankey_data2
sankey_data_o <- sankey_data2
sankey_data_c <- sankey_data2
sankey_data_p <- sankey_data2
sankey_data_d <- sankey_data2

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)   

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 

  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

write_tsv(sankey_data_allTaxa, "data/NEON/all_ind_pavian.txt")
```

```{r}
knitr::include_url("images/sankey_all_ind.txt.html")
```

#### Verrucomicrobiota Coassemblies
```{r}
sankey_data2 <- NEON_MAGs_co %>% 
  filter(Phylum == "Verrucomicrobiota") %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") 


sankey_data2[] <- t(apply(sankey_data2, 1, zoo::na.locf))


sankey_data2 <- sankey_data2 %>% 
    unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")


sankey_data_s <- sankey_data2
sankey_data_g <- sankey_data2
sankey_data_f <- sankey_data2
sankey_data_o <- sankey_data2
sankey_data_c <- sankey_data2
sankey_data_p <- sankey_data2
sankey_data_d <- sankey_data2

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)   

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 

  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

write_tsv(sankey_data_allTaxa, "data/NEON/verruco_co_pavian.txt")
```

```{r}
knitr::include_url("images/sankey_verruco_co.txt.html")
```
#### Verrucomicrobiota Individual
```{r}
sankey_data2 <- NEON_MAGs_ind %>% 
  filter(Phylum == "Verrucomicrobiota") %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") 


sankey_data2[] <- t(apply(sankey_data2, 1, zoo::na.locf))


sankey_data2 <- sankey_data2 %>% 
    unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")


sankey_data_s <- sankey_data2
sankey_data_g <- sankey_data2
sankey_data_f <- sankey_data2
sankey_data_o <- sankey_data2
sankey_data_c <- sankey_data2
sankey_data_p <- sankey_data2
sankey_data_d <- sankey_data2

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)   

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 

  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

write_tsv(sankey_data_allTaxa, "data/NEON/verruco_ind_pavian.txt")
```

```{r}
knitr::include_url("images/sankey_verruco_ind.txt.html")
```


#### Niwot Ridge Individual
```{r}
sankey_data2 <- NEON_MAGs_ind %>% 
  filter(`Site ID` == "NIWO") %>% 
  select(`GTDB-Tk Taxonomy Lineage`) %>% 
  replace_na(list(`GTDB-Tk Taxonomy Lineage` = 'Archaea')) %>% 
  separate(`GTDB-Tk Taxonomy Lineage`, c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species"), ";") 


sankey_data2[] <- t(apply(sankey_data2, 1, zoo::na.locf))


sankey_data2 <- sankey_data2 %>% 
    unite(col = "classification", c(Domain, Phylum, Class, Order, Family, Genus, Species), sep='; ') %>% 
  mutate_at("classification", str_replace, "Archaea", "d__Archaea") %>% 
  mutate_at("classification", str_replace, "Bacteria", "d__Bacteria") %>%  
  mutate_at("classification", str_replace, "; ", "|p__") %>% 
  mutate_at("classification", str_replace, "; ", "|c__") %>% 
  mutate_at("classification", str_replace, "; ", "|o__") %>% 
  mutate_at("classification", str_replace, "; ", "|f__") %>% 
  mutate_at("classification", str_replace, "; ", "|g__") %>% 
  mutate_at("classification", str_replace, "; ", "|s__")


sankey_data_s <- sankey_data2
sankey_data_g <- sankey_data2
sankey_data_f <- sankey_data2
sankey_data_o <- sankey_data2
sankey_data_c <- sankey_data2
sankey_data_p <- sankey_data2
sankey_data_d <- sankey_data2

sankey_data_g$classification <- sub("\\|s__.*", "", sankey_data_g$classification)  
sankey_data_f$classification <- sub("\\|g__.*", "", sankey_data_f$classification)  
sankey_data_o$classification <- sub("\\|f__.*", "", sankey_data_o$classification)  
sankey_data_c$classification <- sub("\\|o__.*", "", sankey_data_c$classification)  
sankey_data_p$classification <- sub("\\|c__.*", "", sankey_data_p$classification)  
sankey_data_d$classification <- sub("\\|p__.*", "", sankey_data_d$classification)   

sankey_data_allTaxa <- bind_rows(sankey_data_s, sankey_data_g, sankey_data_f, sankey_data_o, sankey_data_c, sankey_data_p, sankey_data_d) %>% 
  mutate(classification = as.factor(classification)) %>% 
  count(classification) %>% 

  rename(`#SampleID` = `classification`) %>% 
  rename(`Metaphlan2_Analysis` = `n`) 

write_tsv(sankey_data_allTaxa, "data/NEON/niwot_ind_pavian.txt")
```

```{r}
knitr::include_url("images/sankey_niwot_ind.txt.html")
```


### Verrucomicrobiota Properties

#### Verrucomicrobiota Combined and Individual Assemblies

```{r, fig.height=5, fig.width=7}
ggtree(tree_Verruco, layout="circular")  %<+%
  NEON_data + 
  geom_point2(mapping=aes(color=`Assembly Type`))+
  ggtitle("Verrucomicrobiota Combined and Individual")
```

#### Verrucomicrobiota Genome size and Gene Count

```{r}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Total Number of Bases`, y= `Gene Count` ))+
  geom_point()+
  labs(title = "Verrucomicrobiota Genome Size vs. Gene Count")
```

#### Verrucomicrobiota Order Genome Size Distribution

```{r}
NEON_data_Verruco %>% 
  ggplot(aes(x= Order, y= `Total Number of Bases` ))+
  geom_boxplot()+
  labs(title = "Verrucomicrobiota Order Genome Size Distribution")
```

#### Histogram of Verrucomicrobiota Total Number of Bases

```{r}
NEON_data_Verruco %>% 
ggplot(aes(x = `Total Number of Bases`, color = Phylum)) +
  geom_histogram(bins = 50) 
```

#### Verrucomicrobiota Genus Genome Size Distribution

```{r, fig.width=10}
NEON_data_Verruco %>%
  ggplot(aes(x= Genus, y= `Total Number of Bases` ))+
  geom_boxplot()+
  labs(title = "Verrucomicrobiota Distribution with Total Number of Base")
```

#### Verrucomicrobiota Genera Gene Count Distribution

```{r, fig.width=10}
NEON_data_Verruco %>%
  ggplot(aes(x= Genus, y= `Gene Count` ))+
  geom_boxplot()+
  labs(title = "Verrucomicrobiota Genera Gene Count Distribution")
```

#### 16s rRNA Distribution by Genus

```{r, fig.width=12}
NEON_data_Verruco %>%
  ggplot(aes(x = Genus, y = `16s rRNA`, fill = Genus))+
  geom_boxplot()+
  labs(title = "16s rRNA Distribution by Genus")
```

#### Verrucomicrobiota Genera tRNA Distribution

```{r, fig.width=10}
NEON_data_Verruco %>%
  ggplot(aes(x= Genus, y= `tRNA Genes` ))+
  geom_boxplot()+
  labs(title = "Verrucomicrobiota Genera tRNA Distribution")
```

#### Novel Verrucomicrobiota by Order

```{r}
NEON_data_Verruco %>% 
  filter(is.na(Genus)) %>% 
  ggplot(aes(x= Order))+
  geom_bar()+
  labs(title = "Novel Verrucomicrobiota", y = "Novel Bacterial MAGs", x= "Order") +
  coord_flip()
```

#### Verrucomicrobiota Scaffold Count and Bin Completeness

```{r}
NEON_data_Verruco %>% 
  ggplot(aes(x = `Scaffold Count`, y = `Bin Completeness`)) +
  geom_point() +
  labs(title = "Verrucomicrobiota Scaffold Count and MAG Completeness",
       x = "Scaffold Count",
       y = "Bin Completeness")
```

#### Verrucomicrobiota Histogram of Total Number of Bases at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  filter(Phylum == "Verrucomicrobiota") %>%
ggplot(aes(x = `Total Number of Bases`, color = Phylum)) +
  geom_histogram(bins = 50) 
```

#### Verrucomicrobiota Total Number of Bases and Bin quality

```{r}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Bin Quality`, y= `Total Number of Bases`, color = Phylum))+
  geom_point()+
  labs(title = "Relationship between Total Number of Bases and Bin Quality for Verrucomicrobiota", y = "Total Number of Bases", x= "Bin Quality")
```

### Verrumicrobiota Gene Count and Scaffold Count at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  filter(Phylum == "Verrucomicrobiota") %>%
  ggplot(aes(x= `Scaffold Count`, y= `Total Number of Bases`, color = Phylum ))+
  geom_point()+
  labs(title = "Gene Count and Scaffold Count at NIWO")
```

### Verrucomicrobiota Scaffold Count and Total Number of Bases

```{r}
NEON_data_Verruco %>% 
  ggplot(aes(x= `Scaffold Count`, y= `Total Number of Bases`, color = Phylum ))+
  geom_point()+
  labs(title = "Gene Count and Scaffold Count")
```

### Niwot Ridge

#### Phyla at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  ggplot(aes(x = fct_infreq(Phylum), fill=Class))+
  geom_bar()+
  labs(title = "Phyla at Niwot Ridge", y = "MAGs", x= "Phylum", fill = "Class")+
  coord_flip()
```

#### Genome Size Distribution at Niwot Ridge boxplot

```{r, fig.width=12}
NEON_data_NIWO %>% 
  ggplot(aes(x= Phylum, y= `Total Number of Bases` ))+
  geom_boxplot()+
  labs(title = "Genome Size Distribution at Niwot Ridge")
```

#### Gene Count Distribution at Niwot Ridge

```{r, fig.width=12}
NEON_data_NIWO %>% 
  ggplot(aes(x= Phylum, y= `Gene Count`))+
  geom_boxplot()+
  labs(title = "Gene Count Distribution at Niwot Ridge")
```

#### tRNA Distribution at Niwot Ridge

```{r, fig.width=12}
NEON_data_NIWO %>% 
  ggplot(aes(x= Phylum, y= `tRNA Genes` ))+
  geom_boxplot()+
  labs(title = "tRNA Distribution at Niwot Ridge")
```

#### Novel Bacteria at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  filter(is.na(Genus)) %>% 
  ggplot(aes(x= Phylum))+
  geom_bar()+
  labs(title = "Phyla of Novel MAGs", y = "Novel Bacterial MAGs", x= "Phylum") +
  coord_flip()
```

#### nlcdclass of samples at Niwot Ridge

```{r}
ggplot(NEON_data_NIWO, aes(x = `nlcdClass`, y = `Sample Name`, color = Phylum)) +
  geom_point() +
  labs(title = "Relationship between nlcdclass and samples at NIWO",
       x = "Sample Name",
       y = "nlcdClass")
```

#### Histogram of Total Number of Bases at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
ggplot(aes(x = `Total Number of Bases`, color = Phylum)) +
  geom_histogram(bins = 50) 
```

#### Total Number of Bases and Bin quality at Niwot Ridge

```{r}
NEON_data_NIWO %>% 
  ggplot(aes(x= `Bin Quality`, y= `Total Number of Bases`, color = Phylum))+
  geom_point()+
  labs(title = "Relationship between Total Number of Bases and Bin Quality at NIWO", y = "Total Number of Bases", x= "Bin Quality")
```

### General NEON Properties

#### Total Number of Bases for all NEON samples

```{r}
NEON_data %>% 
  ggplot(aes(x= `Total Number of Bases`))+
  geom_histogram(bins = 50)+
  labs(title="Distribution of Total Number of Bases for all NEON samples", y="MAGs")
```

#### Novel Bacteria per Site

```{r}
NEON_data %>% 
  filter(is.na(Genus)) %>% 
ggplot(aes(x = Site.x))+
  geom_bar() +
  labs(title = "Novel Bacteria per Site", y = "Novel Genera", x= "Site")+
  coord_flip()
```

#### Phyla of Novel MAGs

```{r}
NEON_data %>% 
  filter(is.na(Genus)) %>% 
  ggplot(aes(x= Phylum))+
  geom_bar()+
  labs(title = "Phyla of Novel MAGs", y = "Novel Bacterial MAGs", x= "Phylum") +
  coord_flip()
```


# Discussion

## Environmental Data

We found that, overall, Verrucomicrobiota is a phylum that is found under many different environmental conditions. There were wide spreads of distribution across soil temperature and pH, but the most common conditions were temperatures between 10 and 20 degrees celsius and pH values between 6 and 7. This observation mirrors that of the overall studied MAGs. Verrucomicrobiota were found in 6 different ecosystem subtypes, and these locations were well-distributed throughout their phylogeny. Interestingly, for soil nutrients Verrucomicrobiota tended to be at the high and low ends, but not the middle, of the spectra for both nitrogen and organic carbon amounts, but both of these observations may have been influenced by sample size. Finally, the phylum was found at a variety of different depths, but the most common by far was 0.10 meters (although this is also likely influenced by the types of sampling). Overall, this would imply that Verrucomicrobiota is able to adapt to many different environments, perhaps meaning that they could resist environmental changes.

We also expanded our study of environmental data into different sites, focusing on Niwot Ridge but also examining the other NEON sites due to a lack of data available for Niwot. We found correlations between soil temperature and elevation at Niwot Ridge but not for all sites, as well as a weak one between soil temperature and pH over all NEON sites. The different ecosystem subtypes had different average temperatures which was to be expected. Niwot Ridge was determined to be fully Terrestrial Soil, with a fairly even split between mineral and organic soil samples and a variety of elevations.


## Genome and Assembly Data

In addition to environmental factors we looked into the genomic properties of the Verrucomicrobiota included in this study. There was a clear relationship between genome size and gene count, which was to be expected. Genome size varied slightly between orders of Verrucomicrobiota, but dramatically between genera, with JAFAMB01 having the largest genomes. There was also variety in 16s rRNA counts and tRNA counts. Several novel species of Verrucomicrobiota were discovered, with most of them being of the order Chthoniobacterales For their assemblies, there was not a noticeable relationship between scaffold count, bin completeness, and genome size, and a majority of the samples were of medium bin quality.

At Niwot Ridge, like the total data set, there was a variety of genome sizes and phyla. The most common phylum was Pseudomonadota, and Verrucomicrobiota was the fourth most common type. Of novel bacteria, most were Actinomycetota with Verrucomicrobiota a close second.

Looking at the overall data, there was a wide distribution of genome sizes, but a large majority were less than 5 million bases. Novel bacteria were found at all the sites, but the most common place to find novel species was in the combined assemblies. Of these novel bacteria, most were of the phylum Actinomycetota, and Verrucomicrobiota had an intermediate number. These results suggest that sampling NEON sites, including Niwot Ridge, can provide a variety of information and can contribute new genera and species, which is both exciting and useful for scientific work.


## Project Limitations and Next Steps

This project was significantly limited by a lack of available data for both our phylum and, especially, for our site. For example, in our graph comparing soil temperature and pH at Niwot Ridge, there are only 6 points because very few environmental readings were done at our site. We also were unable to explore other factors that would have been of interest, such as chlorophyll concentration, algal bloom, and other soil nutrients such as sulfur (which is especially relevant to Verrucomicrobiota species). Future studies could investigate these factors as well as following environmental trends over time. All of this could increase our understanding of Verrucomicrobiota and other phyla’s roles in different ecosystems. 


# Conclusion

The study provides a comprehensive insight into the different distributions, environmental preferences, and genomic characteristics of Verrucomicrobiota across various ecosystems and some of the studies having a focus on the Niwot Ridge. Studying the verrucomicrobiota showed a remarkable adaptability to diverse environmental conditions, pH levels, soil depths, and nutrient concentrations. It shows their resilience to different environments and conditions.

Furthermore, the NEON data from the different sites highlight the correlation between some of the factors that are important in ecological research like the soil pH, temperature, and elevation across other NEON sites, emphasizing the need for specific investigations. Analysis of the data showed patterns in genome size and gene count among others.

However, the study was constricted by the data limitations, particularly the lack of other environmental data for the specific site. The data could then be gathered for future research and investigated for the other environmental factors such as chlorophyll concentrations and in particular, their role in sulfur concentrations, advancing our knowledge of microbial communities and ecosystems functioning properly


# Bibliography


